The Kapowski Chronicles: A Complete, Volumetric Cortical Thickness Open Source Pipeline with Evaluation on Public Data

### 2013-02-08 KapowskiChronicles/intro.tex
ENH:  Updating.

% Neuroscientific investigations into cortical morphological
% changes/differences have illuminated interesting correlations with
% normal and pathological neurodevelopment in  addition to cognitive function.
Historically rooted in the meticulous work of von Economo \citep{economo2008},
imaging-based structural analysis of the brain now plays a fundamental role
in identifying the relationship between cortical morphology, disease and cognition.
Quantitative cortical measures are able to identify abnormalities in such conditions as 
Huntington's disease \citep{rosas2002,rosas2005,selemon2004}, 
schizophrenia \citep{nesvag2008}, bipolar disorder \cite{lyoo2006}, Alzheimer's disease and frontotemporal
dementia \citep{du2007,dickerson2009}, Parkinson's disease \citep{jubault2011},
multiple sclerosis \citep{ramasamy2009}, autism \citep{chung2005,hardan2006},
migraines \citep{dasilva2007}, chronic smoking \citep{kuhn2010}, alcoholism \citep{fortier2011},
cocaine addiction \citep{makris2008}, Tourette syndrome in children \citep{sowell2008},
scoliosis in
female adolescents \citep{wang2012}, obsessive compulsive
disorder \citep{shin2007}, ADHD \citep{almeida-montes2012}, obesity \citep{raji2010}, and heritable \citep{peterson2009}
and elderly \citep{ballmaier2004} depression.  Cortical thickness also
varies normally as a function of age \citep{kochunov2011},
gender \citep{luders2006a}, handedness
\citep{luders2006,amunts2007}, intelligence \citep{shaw2006}, athletic
ability \citep{wei2011}, musical ability \citep{bermudez2009,foster2010}, 
tendency toward criminality \citep{raine2011}, and Tetris-playing
ability in female adolescents \citep{haier2009}.  Additionally,
recent studies demonstrate possible functional 
connectivity relationships using cortical thickness measures
\cite{worsley2005,lerch2006,he2007}.
Although these findings
are subject to debate and interpretation \citep{gernsbacher2007}, 
the availability of quantitative
computational methods for extracting such information
has proven invaluable for developing and refining fundamental 
neuroscience hypotheses.

@@

Image volumetric (or meshless) techniques vary both in algorithmic terms as well as
the underlying definition of cortical thickness.  An early, foundational technique is the 
method of \cite{jones2000} in which the inner and outer surface geometry is used to determine the
solution to Laplace's equation in which thickness is measured by integrating along the 
tangents of the resulting field lines spanning the boundary surfaces.  Subsequent contributions
improved upon the original formulation.  For example, in \cite{yezzi2003}, an Eulerian PDE approach
was proposed to facilitate the computation of correspondence paths.  Extending the surface-based
work of \cite{macdonald2000}, the hybrid approach of
\cite{kim2005} uses the discrete Laplacian field to deform the white matter surface mesh towards the 
pial surface.    Although the Laplacian-based approach has several advantages
including generally lower computational times and 
non-crossing correspondence paths, direct correlative assessments with histology
are difficult as the quantified distances 
are not necessarily Euclidean.  Other volumetric algorithms employ coupled
level sets \citep{zeng1999}, model-free intelligent search strategies either normal to 
the gray-white matter interface \citep{scott2009} or using a min-max rule \citep{clement-vachet2011}.
Most relevant to this work is the DiReCT (Diffeomorphic Registration-based 
Cortical Thickness) algorithm proposed by \cite{das2009} in which a 
a registration-based approach proposed where the derived diffeomorphic mapping between the 
white and pial matter surfaces is used to propagate thickness values 
through the cortical gray matter.  A unique benefit of DiReCT is that it
naturally estimates the boundaries of buried sulci by employing a
diffeomorphic constraint on the probabilistic estimate of the gray
matter and cerebrospinal fluid interface.

Despite the variety of techniques for estimating cortical thickness
from imaging data (of which
only a fraction are cited), several common pre-processing components
may be identified.
The most fundamental of these include inhomogeneity correction, skull stripping, and $n$-tissue segmentation 
for differentiating the gray and white matter.  For statistical analysis 
across large populations, construction of population-specific unbiased templates
is also beneficial \cite{}.
In addition, some of these steps might include a crucial registration component, e.g. 
propagating template-based tissue priors for improved segmentation.  The requisite
additional components coupled with the general lack of availability of published
algorithms \cite{kovacevic2006} inhibits performing studies by external researchers 
and makes comparative evaluations difficult.  For example, one recent evaluation 
study \citep{clarkson2011} compared
Freesurfer (a surface-based method) with two volumetric methods \citep{jones2000,das2009}.
Whereas the entire Freesurfer processing pipeline has been made publicly available, 
tuned by the original authors in terms of implementation, and described in great detail 
(specifically in terms of suggested parameters); both volumetric methods were 
implemented solely by the authors of the evaluation (not the actual software developers) using 
unspecified parameters making the comparisons less than
ideal. Further complicating comparisons is distinct processing domains between
volumetric and surface-based techniques and the potential for the introduction
of bias \cite{klein2010}.

In this work, we describe our C$++$ based cortical thickness pipeline
which is freely available as part of the Advanced Normalization Tools
(ANTs) software package.  This includes all the necessary preprocessing steps consisting
of well-vetted previously published algorithms for bias correction \cite{tustison2010},
brain extraction \cite{avants2010a}, $n$-tissue segmentation \cite{avants2011a},
template construction \cite{avants2010}, and image normalization \cite{avants2011}.
We also describe improvements made to the original DiReCT algorithm \cite{das2009}.
Equally as important, we demonstrate how to coordinate
these pipeline components and provide a set of useful parameters
which are employed to analyze the publicly available IXI data set. The
full pipeline and parameter set is 
encapsulated in a well-documented shell script which is also available in ANTs. 
Furthermore, we provide all the derived image data and other processing scripts on 
a github repository specifically meant for this publication.  The
availability of both the code and data permits
the set of results described in this work to be fully reproducible.  This
permits other researchers to contrast their own results against
this baseline processing and to adapt the given volumetric pipeline for measuring
cortical thickness in their own datasets.

>>>>

% Neuroscientific investigations into cortical morphological
% changes/differences have illuminated interesting correlations with
% normal and pathological neurodevelopment in  addition to cognitive function.
Historically rooted in the meticulous work of von Economo \citep{economo2008},
imaging-based structural analysis of the brain plays a fundamental role
in identifying the relationship between cortical morphology, disease and cognition.
Quantitative cortical measures are able to identify abnormalities in such conditions as 
Huntington's disease \citep{rosas2002,rosas2005,selemon2004}, 
schizophrenia \citep{nesvag2008}, bipolar disorder \cite{lyoo2006}, Alzheimer's disease and frontotemporal
dementia \citep{du2007,dickerson2009}, Parkinson's disease \citep{jubault2011}, Williams syndrome \cite{thompson2005},
multiple sclerosis \citep{ramasamy2009}, autism \citep{chung2005,hardan2006},
migraines \citep{dasilva2007}, chronic smoking \citep{kuhn2010}, alcoholism \citep{fortier2011},
cocaine addiction \citep{makris2008}, Tourette syndrome in children \citep{sowell2008},
scoliosis in
female adolescents \citep{wang2012}, obsessive compulsive
disorder \citep{shin2007}, ADHD \citep{almeida-montes2012}, obesity \citep{raji2010}, and heritable \citep{peterson2009}
and elderly \citep{ballmaier2004} depression.  Cortical thickness also
varies normally as a function of age \citep{kochunov2011},
gender \citep{luders2006a}, untreated transsexuality \citep{zubiaurre-elorza2012},  handedness
\citep{luders2006,amunts2007}, intelligence \citep{shaw2006}, athletic
ability \citep{wei2011}, musical ability \citep{bermudez2009,foster2010}, 
tendency toward criminality \citep{raine2011}, and Tetris-playing
ability in female adolescents \citep{haier2009}.  Additionally,
recent studies demonstrate possible functional 
connectivity relationships using cortical thickness measures
\cite{worsley2005,lerch2006,he2007}.
Although these findings
are subject to debate and interpretation \citep{gernsbacher2007}, 
the availability of quantitative
computational methods for extracting such information
has proven invaluable for developing and refining fundamental 
neuroscience hypotheses.

@@

Image volumetric (or meshless) techniques vary both in algorithmic terms as well as
the underlying definition of cortical thickness.  An early, foundational technique is the 
method of \cite{jones2000} in which the inner and outer surface geometry is used to determine the
solution to Laplace's equation where thickness is measured by integrating along the 
tangents of the resulting field lines spanning the boundary surfaces.  Subsequent contributions
improved upon the original formulation.  For example, in \cite{yezzi2003}, an Eulerian PDE approach
was proposed to facilitate the computation of correspondence paths.  Extending the surface-based
work of \cite{macdonald2000}, the hybrid approach of
\cite{kim2005} uses the discrete Laplacian field to deform the white matter surface mesh towards the 
pial surface.    Although the Laplacian-based approach has several advantages
including generally lower computational times and 
non-crossing correspondence paths, direct correlative assessments with histology
are potentially problematic as the quantified distances 
are not necessarily Euclidean.  Other volumetric algorithms employ coupled
level sets \citep{zeng1999}, model-free intelligent search strategies either normal to 
the gray-white matter interface \citep{scott2009} or using a min-max rule \citep{clement-vachet2011}.
Most relevant to this work is the DiReCT (Diffeomorphic Registration-based 
Cortical Thickness) algorithm proposed by \cite{das2009} in which 
a registration-based approach proposed where the derived diffeomorphic mapping between the 
white and pial matter surfaces is used to propagate thickness values 
through the cortical gray matter.  A unique benefit of DiReCT is that it
naturally estimates the boundaries of buried sulci by employing a
diffeomorphic constraint on the probabilistic estimate of the gray
matter and cerebrospinal fluid interface.

Despite the variety of techniques for estimating cortical thickness
from imaging data (of which
only a fraction are cited), several common pre-processing components
may be identified.
The most fundamental of these include inhomogeneity correction, skull stripping, and $n$-tissue segmentation 
for differentiating the gray and white matter.  For statistical analysis 
across large populations, construction of population-specific unbiased templates
is also beneficial \cite{evans2012}.
In addition, some of these steps might include a crucial registration component, e.g. 
propagating template-based tissue priors for improved segmentation.  The requisite
additional components coupled with the general lack of availability of published
algorithms \cite{kovacevic2006} inhibits performing studies by external researchers 
and makes comparative evaluations difficult.  For example, one recent evaluation 
study \citep{clarkson2011} compared
Freesurfer (a surface-based method) with two volumetric methods \citep{jones2000,das2009}.
Whereas the entire Freesurfer processing pipeline has been made publicly available, 
tuned by the original authors in terms of implementation, and described in great detail 
(specifically in terms of suggested parameters); both volumetric methods were 
implemented solely by the authors of the evaluation (not the actual algorithmic developers) using 
unspecified parameters making the comparisons less than
ideal. Further complicating comparisons is distinct processing domains between
volumetric and surface-based techniques and the potential for the introduction
of bias \cite{klein2010}.

In this work, we describe our Insight Toolkit (ITK)-based cortical thickness pipeline
which is freely available as part of the Advanced Normalization Tools
(ANTs) software package.  This includes all the necessary preprocessing steps consisting
of well-vetted previously published algorithms for bias correction \cite{tustison2010},
brain extraction \cite{avants2010a}, $n$-tissue segmentation \cite{avants2011a},
template construction \cite{avants2010}, and image normalization \cite{avants2011}.
We also describe improvements made to the original DiReCT algorithm \cite{das2009}.
Equally as important, we provide explicit coordination between
these pipeline components complete with a set of useful parameters
which are employed to analyze the publicly available IXI data set. The
full pipeline and parameter set is 
encapsulated in a well-documented shell script which is also available in ANTs. 
Furthermore, we provide all the derived image data and other processing scripts on 
the NITRC repository specifically meant for this publication.  The
availability of both the code and data permits
the set of results described in this work to be fully reproducible.  This
permits other researchers to contrast their own results against
this baseline processing and to adapt the given volumetric pipeline for measuring
cortical thickness with their own datasets.

### 2013-02-08 KapowskiChronicles/methods.tex
ENH:  More Brian's edits.
ENH:  Updating.

The ANTs-based cortical thickness estimation workflow is illustrated 
in Figure \ref{fig:pipeline}.  The steps are as follows:
\begin{enumerate}
  \item initial N4 bias correction on input anatomical MRI,
    \textcolor{blue}{do you usually bias correct the raw image?  phil
      had some problems with the automated (otsu) masking employed
      therein .... esp with elderly brain}
  \item brain extraction using a hybrid segmentation/template-based strategy,
  \item alternating between prior-based segmentation and white matter posterior
        probability weighted bias correction,
  \item DiReCT-based cortical thickness estimation, and
  \item optional normalization to specified template.
\end{enumerate}
Each component, including both software and data, is briefly detailed 
below with the relevant references for additional information.

@@

\begin{figure*}
  \centering
  \includegraphics[width=130mm]{Figures/Kapowski_pipeline.pdf}
  \caption{The ANTs T1 processing workflow containing all elements for 
  determining cortical thickness. Not shown is the optional single subject
  to template registration.}
  \label{fig:pipeline}
\end{figure*}

@@

The nonrigid image registration evaluation project is an ongoing 
framework for evaluating image registration algorithms \citep{christensen2006}.
The initial data set introduced into the project consists of 
16 (8 male and 8 female) high resolution skull-stripped brain 
data with 32 cortical labels manually drawn using published protocol.
Given the gray matter labels, the white matter and CSF were identified 
for each of the 16 subjects using Atropos.  Similar to the LPBA40
data set, a NIREP template was created from all 16 subjects and each the
warped labels were used to create probabilistic estimates of the 
labeled region boundaries. These probability maps were used as 
spatial prior probabilities during the 3-tissue segmentation component
of the pipeline.  Using SyN, the NIREP template is registered to the
extracted individual subject brain which is followed by a warping of the 
NIREP priors to the space of the individual subject.  The initial warped 
white matter probability map is used as the weighted confidence mask 
in the follow-up bias correction step.

@@

The IXI data%
\footnote{
http://biomedic.doc.ic.ac.uk/brain-development/
}
consists of approximately 600 healthy subjects imaged at three sites 
using several modalities (T1-weighted, T2-weighted, proton density, magnetic 
resonance angiography, and diffusion tensor imaging).  The 
database also consists of  demographic information such as age, weight,
height, ethnicity, occupation category, educational level, and marital status.
The number of subjects spanning a range of demographic characteristics makes
this a rich data set for validating and exploring correlations with cortical 
thickness measured using the ANTs pipeline.

\begin{figure}
  \centering
  \includegraphics[width=90mm]{Figures/template.pdf}
  \caption{Sample multivariate template constructed from a subset of the IXI data (female, age 40--50).  Axial slices of five of the 37 total subjects from this cohort are shown. }
  \label{fig:template}
\end{figure}

>>>>

The ANTs-based cortical thickness estimation workflow is illustrated 
in Figure \ref{fig:pipeline}.  The steps are as follows:
\begin{enumerate}
  \item initial N4 bias correction on input anatomical MRI,
  \item brain extraction using a hybrid segmentation/template-based strategy,
  \item alternating between prior-based segmentation and white matter posterior
        probability weighted bias correction,
  \item DiReCT-based cortical thickness estimation, and
  \item optional normalization to specified template.
\end{enumerate}
Each component, including both software and data, is briefly detailed 
below with the relevant references for additional information.

@@

\begin{figure*}
  \centering
  \includegraphics[width=170mm]{Figures/Kapowski_pipeline2.pdf}
  \caption{The ANTs T1 processing workflow containing all elements for 
  determining cortical thickness. Not shown is the optional single subject
  to template registration.}
  \label{fig:pipeline}
\end{figure*}

@@

The nonrigid image registration evaluation project is an ongoing 
framework for evaluating image registration algorithms \citep{christensen2006}.
The initial data set introduced into the project consists of 
16 (8 male and 8 female) high resolution skull-stripped brain 
data with 32 cortical labels (cf Table \ref{table:nirep_labels}) manually drawn using a 
published protocol.%
\footnote{
http://www.nirep.org/
}
Given the gray matter labels, the white matter and CSF were identified 
for each of the 16 subjects using Atropos.  Similar to the LPBA40
data set, a NIREP template was created from all 16 subjects and each of the
warped labels were used to create probabilistic estimates of the 
labeled region boundaries. These probability maps were used as 
spatial prior probabilities during the 3-tissue segmentation component
of the pipeline.  Using SyN, the NIREP template is registered to the
extracted individual subject brain which is followed by a warping of the 
NIREP priors to the space of the individual subject.  The initial warped 
white matter probability map is used as the weighted confidence mask 
in the follow-up bias correction step.

\begin{table}
\centering
\begin{tabular*}{0.475\textwidth}{@{\extracolsep{\fill}} l l}
\toprule
  1) L occipital lobe & 2) R occipital lobe \\
  3) L cingulate gyrus & 4) R cingulate gyrus \\
  5) L insula gyrus & 6) R insula gyrus \\
  7) L temporal lobe & 8) R temporal lobe \\
  9) L superior temporal gyr. & 10) R superior temporal gyr. \\
  11) L infero temporal region & 12) R infero temporal region \\
  13) L parahippocampal gyr. & 14) R parahippocampal gyr. \\
  15) L frontal pole & 16) R frontal pole \\
  17) L superior frontal gyrus & 18) R superior frontal gyrus \\
  19) L middle frontal gyrus & 20) R middle frontal gyrus \\
  21) L inferior gyrus & 22) R inferior gyrus \\
  23) L orbital frontal gyrus & 24) R orbital frontal gyrus \\
  25) L precentral gyrus & 26) R precentral gyrus \\
  27) L superior parietal lobule & 28) R superior parietal lobule \\
  29) L inferior parietal lobule & 30) R inferior parietal lobule \\
  31) L postcentral gyrus & 32)   R postcentral gyrus \\  
\bottomrule
\end{tabular*}
\caption{The 32 cortical NIREP labels.}
\label{table:nirep_labels}
\end{table}

@@

The IXI data%
\footnote{
http://biomedic.doc.ic.ac.uk/brain-development/
}
used for the evaluation consists of 577 healthy subjects imaged at three sites 
using several modalities (T1-weighted, T2-weighted, proton density, magnetic 
resonance angiography, and diffusion tensor imaging).  The 
database also consists of  demographic information such as age, weight,
height, ethnicity, occupation category, educational level, and marital status.
The number of subjects spanning a range of demographic characteristics makes
this a rich data set for validating and exploring correlations with cortical 
thickness measured using the ANTs pipeline.

\begin{figure}
  \centering
  \includegraphics[width=90mm]{Figures/template50_60.pdf}
  \caption{Sample multivariate template constructed from a subset of the IXI data (female, age 50--60).  Axial slices of five of the 37 total subjects from this cohort are shown. }
  \label{fig:template}
\end{figure}

### 2013-02-08 KapowskiChronicles/paper.tex
ENH:  Updating.

\usepackage{color}
\usepackage{multirow,booktabs,ctable,array}
\usepackage{lscape}
\usepackage{amsmath}
\usepackage{lineno}
\usepackage{ulem}
\usepackage{setspace}
\usepackage{listings}
\usepackage{float}
\usepackage{listings}
\usepackage{color}

@@

\graphicspath{{./Figures/}
                          }

@@

\begin{abstract} 
Numerous studies have explored the relationship between cortical
structure and brain development, cognitive 
function, and functional connectivity.
The highly convoluted cortical topography makes manual measurements
arduous and often impractical given the population sizes necessary 
for sufficient statistical power.  Computational techniques 
have permitted large-scale studies as they
provide robust and reliable localized measurements characterizing
the cortex with little or no human intervention.  Particularly useful
to the neuroscience community are publicly available tools, such as the popular 
surface-based Freesurfer, which facilitate the testing and refinement of 
hypotheses.  In this paper, we introduce the volume-based 
Advanced Normalization Tools
(ANTs) cortical thickness automated pipeline
comprising well-vetted components such as SyGN (multivariate template construction),
SyN (image registration), N4 (bias
correction),
Atropos ($n$-tissue segmentation), and DiReCT (cortical thickness) 
all developed as part of the ANTs open science effort.  
Complementing the open source aspect of ANTs we confirm 
previous findings of
gender differences, age correlation, and hemispherical asymmetry 
in cortical thickness using the publicly available IXI data
consisting of multimodal structural MRI of approximately 600 subjects.
To further promote open science and use of the proposed tools,
 all scripts and templates used to produce the results
are hosted on the Neuroimaging Informatics Tools and Resources Clearinghouse (NITRC) website.
\end{abstract}

@@

\section{Evaluation}

\section{Discussion}

\section{Conclusions}

>>>>

\usepackage{color}
\usepackage{multirow,booktabs,ctable,array}
\usepackage{lscape}
\usepackage{amsmath}
\usepackage{lineno}
\usepackage{ulem}
\usepackage{setspace}
\usepackage{listings}
\usepackage{float}
\usepackage{listings}
\usepackage{color}
\usepackage{rccol}
\usepackage[table]{xcolor}

@@

\graphicspath{
             {./Figures/}
             {../SPIE2013/KK/IXI_results/} 
             }

@@

\begin{abstract} 
Numerous studies have explored the relationship between cortical
structure and brain development, cognitive 
function and functional connectivity.
The highly convoluted cortical topography makes manual measurements
arduous and often impractical given the population sizes necessary 
for sufficient statistical power.  Computational techniques 
have permitted large-scale studies as they
provide robust and reliable localized measurements characterizing
the cortex with little or no human intervention.  Particularly useful
to the neuroscience community are publicly available tools, such as the popular 
surface-based Freesurfer, which facilitate the testing and refinement of 
hypotheses.  In this paper, we introduce the volume-based 
Advanced Normalization Tools
(ANTs) cortical thickness automated pipeline
comprising well-vetted components such as SyGN (multivariate template construction),
SyN (image registration), N4 (bias
correction),
Atropos ($n$-tissue segmentation), and DiReCT (cortical thickness) 
all developed as part of the ANTs open science effort.  
Complementing the open source aspect of ANTs we confirm 
previous findings of
gender differences, age correlation, and hemispherical asymmetry 
in cortical thickness using the publicly available IXI data
consisting of multimodal structural MRI of approximately 600 subjects.
To further promote open science and use of the proposed tools,
 all scripts and templates used to produce the results
are hosted on the Neuroimaging Informatics Tools and Resources Clearinghouse (NITRC) website.
\end{abstract}

@@

\input{results}

\section{Discussion and Conclusions}
